---
title: "Homework Assignment 1"
author: "Ankit Rao, Credits: Dushyanth Nutulapati"
date: 'Assigned: Oct 24, 2020, Due Sun Nov 01, 2020 11:59PM'
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```
# Reading and Cleaning Data
```{r, cache=TRUE}
nyc<-read.csv("nycflights.csv")

#Cleaning data
str(nyc)

# Factorize columns that need to be a factor
nyc$origin<-as.factor(nyc$origin)
nyc$tailnum<-as.factor(nyc$tailnum)
nyc$month<-as.factor(nyc$month)
nyc$dest<-as.factor(nyc$dest)
nyc$carrier<-as.factor(nyc$carrier)
nyc$flight<-as.factor(nyc$flight)

# Re-check data
str(nyc)
summary(nyc)
```

# Basic Exploration

```{r}
# Using nlevels(nyc$carrier)
```
There are `r nlevels(nyc$carrier)` airlines

```{r}
# Using nrow(nyc[nyc$carrier=="OO",])
```
There are `r nrow(nyc[nyc$carrier=="OO",])` flights with airline code "00"

```{r}
# Using min(nyc$air_time, na.rm = TRUE)
```
The shortest flight out of any NYC airport is `r min(nyc$air_time, na.rm = TRUE)` minutes long
```{r}
# Using nrow(nyc[nyc$carrier=="UA" & nyc$year==2013& nyc$month==1 & nyc$day==13, ])
```
United Airlines on Jan 12th 2013 flew `r nrow(nyc[nyc$carrier=="UA" & nyc$year==2013& nyc$month==1 & nyc$day==13, ])` flights

# Arrival Delay

```{r}
#The average arrival delay in Jan 2013 is
round(mean(nyc[nyc$year==2013 & nyc$month==1, ]$arr_delay, na.rm = TRUE),2) #minutes

# The median arrival delay in Jan 2013 is

round(median(nyc[nyc$year==2013 & nyc$month==1, ]$arr_delay, na.rm = TRUE),2) #minutes

```

Based on the values of mean and median the distribution of arrival delays is right-skewed as mean is significantly higher than median


# Airline Performance

```{r}
# average arrival delays by airline

arr_delay<-tapply(nyc$arr_delay,nyc$carrier,mean,na.rm=TRUE)

barplot(tapply(nyc$arr_delay,nyc$carrier,mean,na.rm=TRUE))

```
THe airline with highest average arrival delay is `r round(max(arr_delay),2)` and the airline with smallest average arrival delay is `r round(min(arr_delay),2)` and there are `r length(arr_delay[arr_delay<0])` airlines with negative average delays namely `r names(arr_delay[arr_delay<0])`


# Air Gain

```{r}
nyc$airgain<-nyc$dep_delay-nyc$arr_delay

airgain_avg<-tapply(nyc$airgain, nyc$carrier, mean, na.rm=TRUE)

barplot(airgain_avg)

```

On average most airlines gain time in the air after departure as we can see `r round(100*length(airgain_avg[airgain_avg>0])/length(airgain_avg),2)`% airlines have positive gains

There are only `r nrow(airgain_avg[airgain_avg<0])` airlines which have negative airgains and they are `r round(airgain_avg[airgain_avg<0],2)` 

# Merging Data Frames

```{r message=FALSE}
library(dplyr)
planes<-read.csv("planes.csv")
# Planes data cleaning
planes$year<-as.factor(planes$year)

merged<-merge(nyc, planes , by.x = "tailnum", by.y = "tailnum")

# Who are the top 5 manufacturers of plane from NYC?
flight_manuf<-merged %>% count(manufacturer)
flight_manuf<-flight_manuf[order(flight_manuf$n,decreasing = TRUE),]
```
Q1) Who are the top 5 manufacturers of plane from NYC?

The top 5 manufactures of all flights going through NYC are

```{r}
top_n(flight_manuf,5)

```


```{r}
# What is the category of seat capacity flowing through NYC?
merged$seats<-as.factor(merged$seats)
nlevels(merged$seats)
flight_seats<-merged %>% count(seats)
flight_seats<-flight_seats[order(flight_seats$n,decreasing = TRUE),]

```
Q2) What is the category of seat capacity flowing through NYC?

The top 3 seating categories of all flights going through NYC are 
```{r}
top_n(flight_seats,3)
```

```{r}
# What is the average airtime across top 3 seating categories of flights?
airtimebycapacity<-aggregate(merged$air_time, list(merged$seats), mean, na.rm=TRUE)
capac_airtime<-airtimebycapacity%>%rename(seat_cat=Group.1,
                           airtime_avg=x)
capacity_merged<-merge(capac_airtime, flight_seats , by.x = "seat_cat", by.y = "seats")
capacity_merged<-capacity_merged[order(capacity_merged$n,decreasing = TRUE),]

```
Q3) What is the average airtime across top 3 seating categories of flights?

The average airtime of top 3 seating categories of all flights going through NYC are 
```{r}
top_n(capacity_merged,3)
```

```{r}
# Which manufacturer has maximum number of fixed wing multi-engine aircrafts
engine_type<-select(merged,type,manufacturer)%>%count(type,manufacturer)
engine_type<-engine_type[order(engine_type$n,decreasing = TRUE),]

```
Q4) Which manufacturer has maximum number of fixed wing multi-engine aircrafts?

The top manufacturer of fixed wing multi-engine aircraft is 

```{r}
top_n(engine_type,1)

```
Q5) What is the average airtime by engine type?


```{r}
airtime_model<-tapply(merged$air_time,merged$type,mean,na.rm=TRUE)

barplot(airtime_model)

```

